---
title: "Ggplot2 Tutorial"
author: "Rick"
date: "01/02/2021"
output: 
  html_document:
  code_folding: "show"
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


```{r load, Libraries}
#Install Tidyverse package
#Install ggthemeassist package
library(tidyverse)
library(ggThemeAssist)
```

We will explore the "ggplot2" package of the tidyverse for data visualisation purposes. The "ggplot2" packages involve the following three mandatory components:

1) Data
2) And aesthetic mapping
3) Geoms (aka objects)

The following components can also be optionally added:

4) Stats (aka transformations)
5) Scales
6) Facets
7) Coordinate systems
8) Position adjustments
9) Themes

Please note that the code in this tutorial was adapted from chapters 3 of the book "R for Data Science" by Hadley Wickham and Garrett Grolemund. The full book can be found at: https://r4ds.had.co.nz/

A good cheatsheet for ggplot2 functions can be found at: https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

Let's create an extremely simple scatterplot. We will use the function 'ggplot()' to do this. 
The format of any ggplot function is this, plus another function to add objects.
The objects in the case of a scatterplot are points. The function we add to it is 'geom_point'. 
These functions rely on a function on the inside called 'aes()'.
The data and aesthetic mapping components can be added to either the 'ggplot()' or geom function.

```{r graph 1}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy))
```
<br>

This is one of the most basic graphs one can make using the ggplot2 frameowork. Next, let's add soem colour.
"geom_point" understands the following aesthetics: x, y, alpha, colour, fill, group, shapre, size and stroke (see help documentation).
Let's map the colour argeument to the variable 'class' from mpg.

```{r graph 2}
ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy, color = class))
```
<br>

This is not the only way to colour objects. Including the color arguement inside the 'aes()' function can map colours to a choice of variable. However, we can also specify colours manually, by specifying colour outside the 'aes()' function. We will also illustrate the 'size', or arguement. 

```{r graph 3}
ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy, size = class), color = "blue")
```
<br>

Lastly, let's examine other objects we can plot using 'ggplot()'. We will create a bar chart using the function 'geom_bar()'. 

```{r graph 4}
ggplot(mpg) +
  geom_bar(aes(x = class))
```

<br>

With the 'geom_bar()' function we have a great use-case for a stat transformation. The following code can be used to transform these counts to proportions. 

```{r graph 5}
ggplot(mpg) +
  geom_bar(aes(x = class, y = stat(prop), group = 1))
```

<br>

Next, let's create a histogram with the 'geom_histogram()' function.

```{r graph 6}
ggplot(mpg) +
  geom_histogram(aes(x = hwy))
```

<br>

The 'geom_histogram()' function accepts the arguement "binwidth", and has two key arguements for color: fill (this controls the overall color), and color (this controls the border).
Let's fill these in:

```{r graph 7}
ggplot(mpg) +
  geom_histogram(aes(x = hwy), binwidth = 5, fill = "navy", color = "gold")
```

<br>

'geom_histogram' provides a great example to modify the scale. Notice in this example the x axis is broken up in units of 10 and does not begin at 0. 
We can modify this with the function 'scale_x_continuous()', as well as the y-axis, with the function 'scale_y_continuous()'. There are three arguements we will feed this function: "breaks", "limits" and "expand". 
"breaks" will define the breaks on an axis, "limits" will define the end of an axis, and the "expand" arguement can be used to start the axis at 0 by using "expand = c(0,0)". 

```{r graph 8}
ggplot(mpg) +
  geom_histogram(aes(x = hwy), binwidth = 5, fill = "navy", color = "gold") +
  scale_x_continuous(breaks = seq(0, 45, 5), limits = c(0, 50), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(0, 90, 10), limits = c(0, 90), expand = c(0,0))
```

<br>

Next, we will create boxplots.

```{r graph 9}
p <- ggplot(mpg) +
  geom_boxplot(aes(x = class, y = cty, fill = class))
p
```

<br>

